Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-30 main.Rmd
minor edits - from section visualization figs

\begin{align*}
\begin{split}
Q_{B,H'}(10) & = 10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert ,\quad since \quad 10 \in T_{1}  \\
  & = 3 \\
\end{split}
\end{align*}
.}
\end{example}

@@

The grammar of graphics introduced a framework to construct statistical graphics by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009pk], which is an alternate and modified parametrization of the grammar suggests that graphics are made up of distinct layers of grammatical elements. Drawing from the grammar of graphics, if <$C_1$, $C_2$, $v$> serves as the basis of visualizing the distribution of the measured variable, the following layers can be specified:

@@

We will discuss the effect of the levels of cyclic granularities on the choice of plots <!--Recall that for cyclic granularities $C_1$ and $C_2$ such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}. $A_i$'s and $B_j$'s are referred to as the levels of $C_1$ and $C_2$ for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ respectively.The levels have an impact on the--> since space and resolution might become a problem if the number of levels are too high. A potential approach could be to categorize the levels as very high/high/medium/low for each cyclic granularity and define some criteria based on usual cognitive power, display size available and the aesthetic mappings. Default values for these categorizations can be chosen based on levels of common temporal granularities like days-of-month, days-of-fortnight and days-of-week.<!--For example, any levels above 31 can be considered as very high, any levels between 14 to 31 can be taken as high and that between 7 to 14 can be taken as medium and below 7 as low.>

@@

If there are levels of cyclic granularities plotted across the  x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarter-of-year. This plot does not work because every quarter does not correspond to all months of the year, for example, the first quarter would not correspond to the month December of an year.

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Distribution of energy consumption displayed through letter value plots. Plot (a) displays consumption across month-of-year faceted by quarter-of-year, (b) across weekday/weekend faceted by quarter-of-year and (c) across quarter-of-year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Analysts should avoid plotting such clashes. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter for weekdays, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}

@@

We will discuss the effect of the mapping of cyclic granularities in this section. When we consider data sets of the form <$C_1$, $C_2$, $v$> with $C_1$ mapped to $x$ position and $C_2$ to facets, then $A_i$'s are placed in close proximity and each $B_j$ represent a group/facet. Gestalt theory suggests that when items are placed in close proximity, people assume that they are in the same group because they are close to one another and apart from other groups. Hence, in this case $A_i$'s are compared against each other within each group.<!-- When $C_2$ is mapped to facets and $C_1$ is mapped to $x$ axis, comparisons would ideally happen across all $B_j$ within each $A_i$--> With the mapping of $C1$ and $C_2$ reversed, emphasis will shift to different behavior of the variables. \autoref{fig:allFig} (b) shows the letter value plot across weekday/weekend faceted by quarter-of-year and \autoref{fig:allFig} (c) shows the same two cyclic granularities with their mapping reversed. \autoref{fig:allFig} (b) helps us to compare weekday and weekend within each quarter and \autoref{fig:allFig} (c) helps to compare quarters within weekend and weekday.

@@

The R package `gravitas` [@R-gravitas] is used to facilitate the systematic exploration here. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing to consider is which cyclic time granularities we can look at exhaustively. Let us consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year). Since the interval of this tsibble is 30 minutes, the temporal granularities may range from half-hour to year. Considering $6$ linear granularities half-hour, hour, day, week, month, year, there could be $(6*5/2) = 15$ (via \ref{sec:chooseharmonies} item #4) circular or quasi-circular granularities that could be formed relating two linear granularities at a time. If these options are considered too numerous, the smallest and largest granularities are good candidates to consider. Hence, we removed half hour and year. Then we will be left $(4*3/2) = 6$ cyclic granularites, namely: "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. Now that we have a list of cyclic granularities to look at, we should be able to compute the multiple-order-up granularities using Section \ref{sec:cyclic-calendar}. <!-- The lubridate package [@Grolemund2011-vm] provides tools to compute single-order-up granularities, \ref{sec:circular-gran-def} (for hour_day, hour_week and day_week) and \ref{sec:quasi-circular-gran-def} (for hour_month, day_month, week_month).  -->
with the index set mapping half-hourly time periods to a set of positive integers.

@@

From the search list, we found six cyclic granularities for which we would like to derive insights of energy behavior. Recalling the data structure <$C1$, $C2$, `general_supply_kwh`> that we would use for exploration, each of these six cyclic granularities can either be mapped to x-axis or to facet. Choosing 2 of the possible 6 granularities, which is equivalent to having ${2 \choose 6} = 30$ candidates for visualization. Fortunately, harmony granularities can be identified among those 30 possibilities to narrow the search. \autoref{tab:harmony-tab} shows 13 harmonies pairs, after removing clashes and any cyclic granularities with levels more than 31, as effective  exploration becomes difficult with many levels (Section \ref{sec:levels}). For each of \autoref{fig:bothcust} (b) and (c), $C1$ is the circular granularity day-type (weekday/weekend) and $C2$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C1$ and $C2$ with $C1$ now denoting hour of the day and $C2$ denoting day-type with distribution geometrically displayed as boxplots.

@@

In \autoref{fig:bothcust} (b), the distribution of energy consumption is plotted across the harmony pair (wknd_wday, hour_day) through an area quantile plot.
The black line is the median, whereas the purple band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely (positive- or right-) skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

Consider looking at violin plots instead of quantile plots to look at the same data in \autoref{fig:bothcust}(c). There is additional information that we can derive looking at the distribution. There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. Also the hours from 7 to 13 look most volatile. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:bothcust}(a), it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicate that looking at different distribution geometry or changing the mapping might shed lights on different aspect of the energy behavior for the same sample population.

@@

However, even if the situation is not periodic and a similar hierarchy can not be formed, it can be interesting to visualize the distribution of a measured variable across relevant cyclic granularities to shed light on the periodic behavior of a non-temporal data set similar to any temporal data set. There are many interesting questions that could possibly be answered with such a data set irrespective of the type of cyclic granularities. <!--We will explore a few interesting questions and understand how the proposed approach in the paper can help answer some of the questions.-->

@@

First, it would be interesting to see if the distribution of total runs vary depending on if a team bats in the first or second inning. The Mumbai Indians (MI) and Chennai Super kings (CSK) <!--are considered one of the best teams in IPL with multiple winning titles and --> appeared in final playoffs from 2010 to 2015. We take their example in order to dive deeper into this question. <!--Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble.--> From Figure \ref{fig:cricex}(a), it can be observed that for the team batting in the first inning there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. This seem to indicate that players feel mounting pressure to score more runs as they approach towards the end of the first inning. Whereas teams batting in the second inning have a set target in mind and are not subjected to such mounting press and may adopt a more conservative strategy, to score runs. Thus winning teams like CSK and MI seem to employ different inning strategies when it comes to their batting order.

@@

```{r cricex, fig.cap= "Runs per over shown with different distribution displays, and granularity arrangements. Plot (a) shows letter value plot across overs faceted by innings. The team batting in the first inning there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. Plot (b) shows box plot of runs per over across an indicator of wickets in previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over.", warning = FALSE, message = FALSE, out.width = "90%"}

>>>>

\begin{align*}
\begin{split}
Q_{B,H'}(10) & = 10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert ,\quad since \quad 10 \in T_{1}  \\
  & = 3 \\
\end{split}
\end{align*}
}
\end{example}

@@

The grammar of graphics introduced a framework to construct statistical graphics by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by @Wickham2009pk, which is an alternate and modified parametrization of the grammar suggests that graphics are made up of distinct layers of grammatical elements. Drawing from the grammar of graphics, if <$C_1$, $C_2$, $v$> serves as the basis of visualizing the distribution of the measured variable, the following layers can be specified:

@@

We will discuss the effect of the levels of cyclic granularities on the choice of plots <!--Recall that for cyclic granularities $C_1$ and $C_2$ such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}. $A_i$'s and $B_j$'s are referred to as the levels of $C_1$ and $C_2$ for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ respectively.The levels have an impact on the--> since space and resolution might become a problem if the number of levels are too high. A potential approach could be to categorize the levels as very high/high/medium/low for each cyclic granularity and define some criteria based on usual cognitive power, display size available and the aesthetic mappings. Default values for these categorizations could be chosen based on levels of common temporal granularities like days of the month, days of the fortnight or days of the week.<!--For example, any levels above 31 can be considered as very high, any levels between 14 to 31 can be taken as high and that between 7 to 14 can be taken as medium and below 7 as low.>

@@

If there are levels of cyclic granularities plotted across the  x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarters of the year. This plot does not work because every quarter does not correspond to all months of the year, for example, the first quarter of the year would never correspond to the month December of an year.

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Distribution of energy consumption displayed through letter value plots. Plot (a) displays consumption across month-of-year faceted by quarters of the year, (b) across weekday/weekend faceted by quarters of the year and (c) across quarters of the year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Analysts should avoid plotting such clashes. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter for weekdays, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}

@@

We will discuss the effect of the mapping of cyclic granularities in this section. When we consider data sets of the form <$C_1$, $C_2$, $v$> with $C_1$ mapped to $x$ position and $C_2$ to facets, then $A_i$'s are placed in close proximity and each $B_j$ represent a group/facet. Gestalt theory suggests that when items are placed in close proximity, people assume that they are in the same group because they are close to one another and apart from other groups. Hence, in this case $A_i$'s are compared against each other within each group.<!-- When $C_2$ is mapped to facets and $C_1$ is mapped to $x$ axis, comparisons would ideally happen across all $B_j$ within each $A_i$--> With the mapping of $C1$ and $C_2$ reversed, emphasis will shift to different behavior of the variables. \autoref{fig:allFig} (b) shows the letter value plot across weekday/weekend faceted by quarters of the year and \autoref{fig:allFig} (c) shows the same two cyclic granularities with their mapping reversed. \autoref{fig:allFig} (b) helps us to compare weekday and weekend within each quarter and \autoref{fig:allFig} (c) helps to compare quarters within weekend and weekday.

@@

The tsibble object `smart_meter10` from R package `gravitas` [@R-gravitas] is used to facilitate the systematic exploration here. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing to consider is which cyclic time granularities we can look at exhaustively. Let us consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year). Since the interval of this tsibble is 30 minutes, the temporal granularities may range from half-hour to year. Considering $6$ linear granularities half-hour, hour, day, week, month, year, there could be $(6*5/2) = 15$ (via \ref{sec:data-structure}) circular or quasi-circular granularities that could be formed relating two linear granularities at a time. If these options are considered too numerous, the smallest and largest granularities are good candidates to consider. Hence, we removed half hour and year. Then we will be left $(4*3/2) = 6$ cyclic granularites, namely: "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. Now that we have a list of cyclic granularities to look at, we should be able to compute the multiple-order-up granularities using Section \ref{sec:cyclic-calendar}. <!-- The lubridate package [@Grolemund2011-vm] provides tools to compute single-order-up granularities, \ref{sec:circular-gran-def} (for hour_day, hour_week and day_week) and \ref{sec:quasi-circular-gran-def} (for hour_month, day_month, week_month).  -->
with the index set mapping half-hourly time periods to a set of positive integers.

@@

From the search list, we found six cyclic granularities for which we would like to derive insights of energy behavior. Recalling the data structure <$C1$, $C2$, `general_supply_kwh`> that we would use for exploration, each of these six cyclic granularities can either be mapped to x-axis or to facet. Choosing 2 of the possible 6 granularities, which is equivalent to having ${6 \choose 2} = 30$ candidates for visualization. Fortunately, harmony granularities can be identified among those 30 possibilities to narrow the search. \autoref{tab:harmony-tab} shows 13 harmonies pairs, after removing clashes and any cyclic granularities with levels more than 31, as effective  exploration becomes difficult with many levels (Section \ref{sec:levels}). For each of \autoref{fig:bothcust} (b) and (c), $C1$ is the circular granularity day-type (weekday/weekend) and $C2$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C1$ and $C2$ with $C1$ now denoting hour of the day and $C2$ denoting day-type with distribution geometrically displayed as boxplots.

@@

In \autoref{fig:bothcust} (b), the distribution of energy consumption is plotted across the harmony pair (weekday/weekend, hour of the day) through an area quantile plot. The black line is the median, whereas the purple band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely (positive- or right-) skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

Consider looking at violin plots instead of quantile plots to look at the same data in \autoref{fig:bothcust}(c). There is additional information that we can derive looking at the distribution. There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. <!--Also the hours from 7 to 13 look most volatile.--> If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:bothcust}(a), it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicate that looking at different distribution geometry or changing the mapping might shed lights on different aspect of the energy behavior for the same sample population.

@@

However, even if the situation is not periodic and a similar hierarchy can not be formed, it can be interesting to visualize the distribution of a measured variable across relevant cyclic granularities to shed light on the aperiodic behavior of a non-temporal data set similar to aperiodic events like formal meetings, workshops, conferences, school semesters in a temporal set up. There are many interesting questions that could possibly be answered with such a data set irrespective of the type of cyclic granularities. <!--We will explore a few interesting questions and understand how the proposed approach in the paper can help answer some of the questions.-->

@@

First, it would be interesting to see if the distribution of total runs vary depending on if a team bats in the first or second inning. The Mumbai Indians (MI) and Chennai Super kings (CSK) <!--are considered one of the best teams in IPL with multiple winning titles and --> appeared in final playoffs from 2010 to 2015. We take their example in order to dive deeper into this question. <!--Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble.--> From Figure \ref{fig:cricex}(a), it can be observed that for the team batting in the first inning there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. This seem to indicate that players feel mounting pressure to score more runs as they approach towards the end of the first inning. Whereas teams batting in the second inning have a set target in mind and are not subjected to such mounting pressure and may adopt a more conservative strategy, to score runs. Thus winning teams like CSK and MI seem to employ different inning strategies when it comes to their batting order.

@@

```{r cricex, fig.cap= "Runs per over shown with different distribution displays, and granularities. Plot (a) shows letter value plot across overs faceted by innings. For the team batting in the first inning there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. Plot (b) shows box plot of runs per over across an indicator of wickets in previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over.", warning = FALSE, message = FALSE, out.width = "90%"}
